Results

### 2025-09-03 notebooks/02_stats.qmd
feat: add plasma beta parameters

taus = Second.(2 .^ (1:6))
# taus = Second.(2 .^ (0:6))
df = workload(taus)
@chain df begin
    DC.compute_Alfvenicity_params!()
    DC.compute_params!()
    @transform!(:B_n_mva_norm = abs.(:B_n_mva ./ :B_mag))
    DC.classify!(:B_n_mva_norm)
end

# Count the number of events for each type
df_event_counts = @chain df begin
    groupby(:type)
    combine(nrow)
end
```

@@

```{julia}
subset_𝐧(df, 𝐧) = 𝐧 == :mva ? filter_low_mva_quality(df) : df
fname = "properties_hist"

@@

##

Joint distributions of thickness and current density

@@

    tdf = @chain df begin
        @subset(.!isnan.(:B_n_mva_norm) .& .!isnan.(:σ_c); view=true)
        dropmissing([:B_n_mva_norm, :duration, :σ_c])
        @transform!(:duration = Float64.(ustrip.(:duration .|> u"s")))
        subset_𝐧(𝐧)
    end

    plt = data(tdf) * mapping(row=:id, col=:enc)
    vis = visual(Scatter; markersize=6)
    fig = Figure(; size=(1200, 800))
    axis = (;)

@@

    B_n_spec = plt * mapping(:B_n_mva_norm, :ω_in) * vis
    ΔV_spec = plt * mapping(:ΔV_ratio, :ΔV_cosθ) * (vis + AoG.smooth() * ΔV_cosθ_map)
    σc_σr_spec = plt * mapping(:σ_c, :σ_r) * (vis + AoG.smooth() * visual(color=:red))

    draw!(fig[1, 1], B_n_spec; axis)
    draw!(fig[1, 2], ΔV_spec; axis=(; limits=((0, 2), (-1, 1),)))
    draw!(fig[2, 1], σc_σr_spec)

@@

Joint distributions of Q_sonnerup and other variables

```{julia}
CairoMakie.activate!()

view = true
𝒎 = var_mapping(; 𝐧)

@@

tdf = @chain df begin
    dropmissing(vars)
    @rsubset(all(!isnan, (:B_n_mva_norm, :Q_sonnerup, :σ_c)); view=true)
    @transform!(
        :duration = Float64.(ustrip.(:duration .|> u"s")),
        :B_n_mva_norm = Float64.(:B_n_mva_norm),
    )
end

@@

@@

layout = mapping(row=:id, col=:enc)
plt = data(tdf) * layout
vis = visual(Scatter; markersize=6)

@@

Bn_Q_spec = data(Bn_df) * mapping(:B_n_mva_norm, :Q_sonnerup) * layout * (vis + AoG.linear() * Q_group * line_vis)
duration_Q_spec = plt * mapping(:duration, :Q_sonnerup) * (vis + AoG.smooth() * Q_group * line_vis)
σc_Q_spec = plt * mapping(:σ_c, :Q_sonnerup) * (vis + AoG.linear() * Q_group * line_vis)
σr_Q_spec = plt * mapping(:σ_r, :Q_sonnerup) * (vis + AoG.linear() * Q_group * line_vis)
A_He_Q_spec = data(A_He_df) * mapping(:A_He, :Q_sonnerup => abs) * layout * (vis + AoG.smooth() * A_He_group * line_vis)

draw!(fig[1, 1], Bn_Q_spec)
draw!(fig[1, 2], duration_Q_spec; axis=(; xscale=log10))
draw!(fig[2, 1], σc_Q_spec)
draw!(fig[2, 2], σr_Q_spec)
draw!(fig[3, 1], A_He_Q_spec)

# add_labels!([ga, gb, gc])
# easy_save("Q_sonnerup_joint_dist"; formats, force=true)
display(fig)
```

@@

# Overall correlation matrix
println("Overall correlation matrix:")
cor_vars = [:Q_sonnerup, :duration, :B_n_mva_norm, :σ_c, :σ_r, :L_n_mva_norm, :J_m_max_mva_norm]
overall_corr = nancor(Matrix(df_corr[!, cor_vars]))
display(overall_corr)
# heatmap(cor_vars, cor_vars, overall_corr)
# pairplot(df_corr[!, cor_vars])

# Group by different missions and compute correlations
println("\nCorrelation analysis by mission:")
mission_stats = @chain df_corr begin
    groupby(:id)
    @combine(
        :count = nrow,
        :corr_Q_duration = nancor(:Q_sonnerup, :duration),
        :corr_Bn_duration = nancor(:B_n_mva_norm, :duration),
        :corr_Q_Bn = nancor(:Q_sonnerup, :B_n_mva_norm),
        :mean_Q = nanmean(abs.(:Q_sonnerup)),
        :mean_duration = nanmean(:duration),
        :mean_Bn = nanmean(:B_n_mva_norm)
    )
end

display(mission_stats)

# # Statistical significance tests
# test_q = CorrelationTest(df_corr.duration_s, df_corr.Q_sonnerup)
# test_vl = CorrelationTest(df_corr.duration_s, collect(skipmissing(df_corr.V_l_ratio)))

>>>>

taus = Second.(2 .^ (1:6))
# taus = Second.(2 .^ (0:6))
df = workload(taus)

_float64(x) = Float64(x)
_float64(::Missing) = missing

@chain df begin
    DC.compute_Alfvenicity_params!()
    DC.compute_params!()
    @transform!(
        :B_n_mva_norm = abs.(:B_n_mva ./ :B_mag),
        :duration = _float64.(ustrip.(:duration .|> u"s"))
    )
    DC.classify!(:B_n_mva_norm)
end

# Count the number of events for each type
df_event_counts = @chain df begin
    groupby(:type)
    combine(nrow)
end

subset_𝐧(df, 𝐧) = 𝐧 == :mva ? filter_low_mva_quality(df) : df
```

## Associations and independence test

```{julia}

println("\n" * "="^60)
println("ASSOCIATION ANALYSIS FOR ALFVÉNICITY")
println("="^60)

# Define the variable pairs you want to analyze with transformations embedded
alfvenicity_pairs = [
    (:duration, :Q_sonnerup => abs, "Duration vs Alfvénicity (Q_sonnerup)"),
    (:duration, :V_l_ratio, "Duration vs Velocity Ratio (V_l_ratio)"),
    (:B_n_mva_norm, :Q_sonnerup => abs, "Magnetic Field Normal vs Alfvénicity"),
    (:σ_c, :Q_sonnerup, "Cross helicity vs Alfvénicity"),
    (:σ_r, :Q_sonnerup),
    (:β, :Q_sonnerup => abs),
    (:A_He => x -> ismissing(x) ? missing : only(x), :Q_sonnerup => abs, "Helium Abundance vs Alfvénicity"),
    (:J_m_max_mva_norm, :Q_sonnerup => abs, "Current density vs Alfvénicity"),
    (:L_n_mva_norm, :Q_sonnerup => abs, "Thickness vs Alfvénicity"),
    (:A_He => x -> ismissing(x) ? missing : only(x), :J_m_max_mva_norm),
    (:A_He => x -> ismissing(x) ? missing : only(x), :L_n_mva_norm),
    (:β, :J_m_max_mva_norm),
    (:β, :L_n_mva_norm),
    (:β, :A_He),
]

# Run overall analysis
println("OVERALL ASSOCIATION ANALYSIS")
println("="^50)
analyze_associations(df, alfvenicity_pairs)

# Grouped analysis
# println("GROUPED ASSOCIATION ANALYSIS (by $(group_by))")
println("="^50)
analyze_associations(df, alfvenicity_pairs; group_by=:id)
```

```md
Association Measure Interpretation:
Values range from 0 (no association) to 1 (perfect association):
- 0.0 - 0.1: Very weak association
- 0.1 - 0.3: Weak association  
- 0.3 - 0.5: Moderate association
- 0.5 - 0.7: Strong association
- 0.7 - 1.0: Very strong association

Association Methods Used:
1. Distance Correlation: Non-parametric measure detecting linear & nonlinear dependencies
2. Pearson Correlation: Standard linear correlation coefficient  
3. Spearman Correlation: Rank-based correlation (monotonic relationships)
4. Chatterjee Correlation: New robust measure for general dependencies
5. Mutual Information: Information-theoretic measure of shared information

Note: These are association strengths, not statistical significance tests.
For independence testing, use with caution - consider both effect size and sample size.
```

[](../association_results.txt)

```md

🔎 Step 1. General observations
	•	Cross helicity vs Alfvénicity absolutely dominates:
	•	Pearson = 0.84 overall (0.92 in PSP, 0.76 in Wind).
	•	Chatterjee = 0.47–0.53.
	•	MI = 0.58–0.97.
→ This is a very strong dependence by any metric, which is expected since Alfvénicity is conceptually related to cross helicity.
	•	Magnetic field normal vs Alfvénicity shows weak-to-moderate positive linear correlation (~0.17–0.19), but Chatterjee and MI stay low. Suggests a weak linear trend, not a strong monotonic or nonlinear relationship.
	•	Σ_r vs Q_sonnerup:
	•	Small negative Pearson (~ -0.1).
	•	But Chatterjee 0.09–0.15 and MI up to 0.24 in PSP → weak, possibly nonlinear dependence.
	•	This might indicate subtle structure missed by linear correlation.
	•	Duration, velocity ratio, thickness, helium abundance, current density:
	•	All show weak or negligible correlations (|Pearson| < 0.25, Chatterjee mostly < 0.06, MI close to 0).
	•	Suggests near independence, with only minor dataset-specific effects (e.g., PSP helium abundance at -0.24).

⸻

🔎 Step 2. Dataset comparison (PSP vs Wind)
	•	Cross helicity vs Alfvénicity is strong in both spacecraft, but even stronger in PSP (r ~ 0.92 vs 0.76).
	•	Magnetic field normal vs Alfvénicity is similar across PSP/Wind (r ~ 0.17–0.18).
	•	Σ_r vs Q_sonnerup is stronger in PSP (MI ~0.24 vs 0.05 in Wind).
	•	Helium abundance vs Alfvénicity:
	•	PSP: moderate negative Pearson (-0.24).
	•	Wind: near zero.
→ Suggests PSP-specific dependence (possibly tied to solar wind composition closer to the Sun).
	•	The rest (duration, velocity ratio, thickness, current density) are consistently weak in both datasets.

⸻

🔎 Step 3. Practical interpretation
	1.	Cross helicity is the defining feature of Alfvénicity.
The high correlations confirm that your Alfvénicity measure (Q_sonnerup) essentially tracks cross helicity. This is expected, but the strength of correlation (0.8–0.9) validates consistency across datasets.
	2.	Other plasma/structural properties are only weakly linked.
Magnetic field orientation (normal) shows a weak positive association, but not much more.
	3.	Σ_r (residual energy measure?) has a modest but interesting nonlinear association with Alfvénicity, especially in PSP. Could hint at a subtle dependence worth exploring with scatter plots or conditional analysis.
	4.	Helium abundance only matters in PSP.
The -0.24 Pearson suggests Alfvénicity tends to be lower in He-rich regions, but this trend does not appear in Wind. That could reflect heliocentric distance effects.

⸻

✅ Summary of main findings:
	•	Strong dependence: Cross helicity ↔ Alfvénicity (core relationship).
	•	Weak but consistent: Magnetic field normal, Σ_r (especially in PSP).
	•	Dataset-specific: Helium abundance (negative in PSP, absent in Wind).
	•	Effectively independent: Duration, velocity ratio, thickness, current density.

⸻
```

Would you like me to help you visualize these dependencies (scatter plots, density plots, MI heatmaps) to confirm the weak vs strong associations, or do you just need a written interpretation for now?

@@

```{julia}
fname = "properties_hist"

@@

## Joint distributions

### Thickness and current density

@@

    tdf = @chain df begin
        @rsubset(all(!isnan, (:B_n_mva_norm, :Q_sonnerup, :σ_c, :ΔV_ratio, :ΔV_cosθ)); view=true)
        dropmissing([:B_n_mva_norm, :duration, :σ_c])
        @transform!(:duration = Float64.(:duration))
        subset_𝐧(𝐧)
    end

    β_A_He_df = @chain df begin
        @rsubset(all(!isnan, (:β, :A_He)); view=true)
        dropmissing([:β, :A_He])
        @transform!(:log_β = log10.(:β))
        @rsubset(:A_He < 20, :log_β > -1.5)
    end

    layout = mapping(row=:id, col=:enc)

    plt = data(tdf) * layout
    vis = visual(Scatter; markersize=6)
    fig = Figure(; size=(1200, 800))
    axis = (;)

@@

    B_n_spec = plt * mapping(:B_n_mva_norm, :ω_in) * vis
    ΔV_spec = plt * mapping(:ΔV_ratio, :ΔV_cosθ) * (vis + AoG.linear() * ΔV_cosθ_map)
    σc_σr_spec = plt * mapping(:σ_c, :σ_r) * (vis + AoG.smooth() * visual(color=:red))

    A_He_group = mapping(group=:A_He => >(1.5), color=:A_He => >(1.5))

    β_A_He_spec = data(β_A_He_df) * layout * mapping(:log_β, :A_He) * (vis + AoG.smooth(; degree=1) * A_He_group)

    draw!(fig[1, 1], B_n_spec; axis)
    draw!(fig[1, 2], ΔV_spec; axis=(; limits=((0, 2), (-1, 1),)))
    draw!(fig[2, 1], σc_σr_spec)
    draw!(fig[2, 2], β_A_He_spec)

@@

### Q_sonnerup and other variables

```{julia}
view = true
𝒎 = var_mapping(; 𝐧)

@@

tdf = @chain df begin
    dropmissing(vars)
    @rsubset(all(!isnan, (:B_n_mva_norm, :Q_sonnerup, :σ_c)); view=true)
    @transform!(
        :duration = Float64.(:duration),
        :B_n_mva_norm = Float64.(:B_n_mva_norm),
    )
end

@@

β_df = @chain df begin
    @rsubset(all(!isnan, [:β, :Q_sonnerup]); view=true)
    dropmissing([:β, :Q_sonnerup])
    @transform!(:log_β = log10.(:β))
end

@@

layout = mapping(row=:id, col=:enc)
plt = data(tdf) * layout
vis = visual(Scatter; markersize=5)

@@

Bn_Q_spec = data(Bn_df) * mapping(:B_n_mva_norm, :Q_sonnerup => abs) * layout * (vis + AoG.linear() * line_vis * mapping(color=:id))
duration_Q_spec = plt * mapping(:duration, :Q_sonnerup) * (vis + AoG.smooth() * Q_group * line_vis)
σc_Q_spec = plt * mapping(:σ_c, :Q_sonnerup) * (vis + AoG.linear() * Q_group * line_vis)
σr_Q_spec = plt * mapping(:σ_r, :Q_sonnerup) * (vis + AoG.linear() * Q_group * line_vis)
A_He_Q_spec = data(A_He_df) * mapping(:A_He, :Q_sonnerup => abs) * layout * (vis + AoG.smooth() * A_He_group * line_vis)
β_Q_spec = data(β_df) * mapping(:log_β, :Q_sonnerup => abs) * layout * (vis + AoG.smooth(; degree=4) * line_vis * mapping(color=:id))

draw!(fig[1, 1], Bn_Q_spec)
draw!(fig[1, 2], duration_Q_spec; axis=(; xscale=log10))
draw!(fig[2, 1], σc_Q_spec)
draw!(fig[2, 2], σr_Q_spec)
draw!(fig[3, 1], A_He_Q_spec)
draw!(fig[3, 2], β_Q_spec)

# add_labels!([ga, gb, gc])
easy_save("Q_sonnerup_joint_dist"; formats, force=true)
display(fig)
```

@@

# Overall correlation matrix
println("Overall correlation matrix:")
cor_vars = [:Q_sonnerup, :duration, :B_n_mva_norm, :σ_c, :σ_r, :L_n_mva_norm, :J_m_max_mva_norm]
_cor_func = df -> nancor(Matrix(df[!, cor_vars]))
overall_corr = _cor_func(df_corr)
display(overall_corr)
# # Statistical significance tests
# test_q = CorrelationTest(df_corr.duration_s, df_corr.Q_sonnerup)
# test_vl = CorrelationTest(df_corr.duration_s, collect(skipmissing(df_corr.V_l_ratio)))
